---
title: "Assignment 5: Data Visualization"
author: "Iman Byndloss"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1: Used library() to load relevant packages. 
# Used read.csv() to assign each set of raw data as a dataset in RStudio.
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)

Nutrients <- read.csv(
  here(
    "Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
  )
Litter <- read.csv(
  here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv")
  )

#2: # Used class() to determine the class for "sampledate" and "collectDate" in the Nutrients and Litter dataframes. 
# Used as.Date() to establish format for the "Date" column of each date frame.
# Ensured it was saved to the appropriate place with "<-".
class(Nutrients$sampledate)
Nutrients$sampledate <- as.Date(Nutrients$sampledate, format = "%Y-%m-%d")

class(Litter$collectDate)
Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d")
```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3: Looked up help(theme).
# Used theme_base() + theme() to establish a custom theme.
custom_grid <- theme_base() + 
  theme(
    panel.background = element_rect(fill="white"), 
    panel.grid.major = element_line(color="gray", linewidth = 0.4),
    panel.grid.minor = element_line(color="gray", linewidth = 0.2), # Set a grey background with white grid lines
    plot.title = element_text(size=10, face="bold",hjust=0.5), # Set text size, emboldened it, and centered plot title
    axis.title = element_text(size=8), # Set text size for axis title
    axis.ticks = element_line(color="black"),
    axis.ticks.length=unit(0.15,"cm"), # Set length of axis ticks
    legend.box.background = element_rect(color="black", size=0.5), # Set a black border around legend
    strip.text = element_text(face="bold", size=8), # Set the facet labels to bold small text
    legend.position = "right" # Set legend position to right of graphs
    )
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4: Used
Nutrients %>%
  ggplot(aes(x=tp_ug, y=po4)) +
  geom_point() +
  facet_wrap(facets=vars(lakename), nrow = 2) +
  geom_smooth(method=lm, se=FALSE) + # se=FALSE removes shaded confidence interval
  labs(
    title="Total phsophorus versus phosphate for Peter and Paul lakes",
    x="Total Phosphorus",
    y="Phosphate") +
  xlim(-4, 150) + # Adjusted to remove extra black space from x axis
  ylim(0, 50) + # Adjusted to hide extreme value above 300 on y axis
  custom_grid
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5: Prior to graphing by month, converted months to abbreviations.
# Then, used ggplot() to creat each of the three plots independently.
# Finally, used cowplot's plot_grid() to combine the plots.
Nutrients$month <- factor(Nutrients$month, levels=1:12, labels=month.abb)

Temp <- 
  ggplot(Nutrients, aes(x = month, y = temperature_C, color = lakename)) +
  geom_boxplot() +
  labs(
    title="Temperature per month for Peter and Paul lakes",
    x="Month", 
    y="Temperature (°C)",
    color="Lake Name") +
  scale_x_discrete(limits = month.abb[5:11]) + 
  # Set limit for x axis from May to Nov (the months with data for temp)
  custom_grid
print(Temp)
  
TP <- 
  ggplot(Nutrients, aes(x = month, y = tp_ug, color = lakename)) +
  geom_boxplot() +
  labs(
    title="Total phosphorous per month for Peter and Paul lakes",
    x="Month", 
    y="Total Phosphorous",
    color="Lake Name") +
  scale_x_discrete(limits = month.abb[5:11]) + 
  ## Set limit for x axis from May to Nov (the months with data for temp for the sake of combining in the next step)
  custom_grid
print(TP)

TN <-
  ggplot(Nutrients, aes(x = month, y = tn_ug, color = lakename)) +
  geom_boxplot() +
  labs(
    title="Temperature per month for Peter and Paul lakes",
    x="Month", 
    y="Total Nitrogen",
    color="Lake Name") +
  scale_x_discrete(limits = month.abb[5:11]) + 
  # Set limit for x axis from May to Nov (the months with data for temp for the sake of combining in the next step)
  custom_grid
print(TN)

library(cowplot)
plot_grid(
  Temp + theme(axis.title.x = element_blank(), legend.position = "none"), # Removed x axis title and legend
  TP + theme(axis.title.x = element_blank(), legend.position="none"), # Removed x axis title and legend 
  TN, 
  nrow = 3, # nrow specifies the number of rows
  align = 'v', # align ensures it is aligned vertically 'v' as opposed to horizontally 'h'
  rel_heights = c(0.95, 1.15, 1.5)) # rel_heights establishes the relative heights of the plots; chosen based on size of y axis
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6


#7


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: