---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Student Name"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1 
# Used getwd() to check the working directory.
getwd()

# Used library() to load relevant packages.
library(tidyverse)
library(agricolae)
library(here)
library(ggthemes)

# Used read.csv() and here() to import specified NTL-LTER raw data file.
LakeChemistry <- read.csv(here(
  "Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), 
  stringsAsFactors = TRUE)

# Used as.Date() to establish sampledate as a date with the format %m/%d/%Y.
# The four digit year corresponds to "%Y", not "%y".
LakeChemistry$sampledate <- as.Date(LakeChemistry$sampledate, format="%m/%d/%Y")

#2
# COPIED FROM A05: Used theme_base() + theme() to establish a custom theme.
custom_grid <- theme_base() + 
  theme(
    panel.background = element_rect(fill="white"), 
    panel.grid.major = element_line(color="gray", linewidth = 0.4),
    panel.grid.minor = element_line(color="gray", linewidth = 0.2), 
    # Set a grey background with white grid lines
    plot.title = element_text(size=10, face="bold",hjust=0.5), 
    # Set text size, emboldened, and centered for plot title
    axis.title = element_text(size=8), 
    # Set text size for axis title
    axis.ticks = element_line(color="black"),
    axis.ticks.length=unit(0.15,"cm"), 
    # Set length of axis ticks
    legend.box.background = element_rect(color="black", size=0.5), 
    # Set a black border around legend
    strip.text = element_text(face="bold", size=8), 
    # Set the facet labels to bold small text
    legend.position = "right" 
    # Set legend position to right of graphs
    )

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The mean lake temperature recorded during July does not change significantly with depth across all lakes.
Ha: The mean lake temperature recorded during July changes significantly with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 째C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
# Extracted month from "sampledate" in preparation for filtering.
LakeChemistry$month <- month(LakeChemistry$sampledate)

# Converted month to a factor with 12 levels, labelled with month names.
LakeChemistry$month <- factor(LakeChemistry$month, 
                                         levels=1:12, 
                                         labels=month.abb)

# Filtered the dataframe as suggested, using %>% (pipe function).
FilteredLakeChemistry <- LakeChemistry %>%
  filter(month=="Jul") %>%
  select(lakename,year4,daynum,depth,temperature_C) %>%
  na.omit() # na.omit() removes all NA values while drop_na() specified columns

#5
# Based on suggested parameters,
# created a scatter plot of temp vs depth with multiple functions.
tempVSdepth <- FilteredLakeChemistry %>%
  ggplot(aes(x=depth,y=temperature_C)) +
  geom_point() +
  geom_smooth(method="lm") +
  ylim(0,35) +
  labs(x="Depth (m)", y="Temperature (째C)") +
  custom_grid
tempVSdepth

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: Based on the figure, as depth increases, temperature decreases; in other words, there is an inverse relationship between these two variables. The distribution of points suggests that the linearity of this trend is not consistent for all depths, with the deepest parts of the lakes from ~10-15m displaying similar temperatures.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
# Fit a simple linear regression to the two variables with lm().
# Used summary() to produce a result summary.
tempVSdepth.regression <- 
  lm(FilteredLakeChemistry$temperature_C ~ 
       FilteredLakeChemistry$depth)
summary(tempVSdepth.regression)
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: According to the linear regression model, there is a statitically significant inverse relationship between temperature and depth; for every 1 meter increase in depth, there is an approximate 1.95째C decrease in temperature (Estimate = -1.95, t value = -165.8, p value = <2e-16). Depth explains 74% of variablilty in temperature with 9726 degrees of freedom (Adjusted R-squared=0.7387). Ultimately, the null hypothesis is rejected in favor of the alternative hypothesis.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
temp.ALL.regression <- lm(data=FilteredLakeChemistry, temperature_C ~ year4 + daynum + depth)
step(temp.ALL.regression)

#10
summary(temp.ALL.regression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: With the AIC being smallest with all three variables (AIC=26065.53), the AIC method suggests "year4", "daynum", and "depth" all be included as explanatory variables to predict temperature in a multiple regression. This model explains 74.11% of the observed variance (Adjusted R-squared=0.7411), which is a slight improvement from the previous model that only contained depth as an explanatory variable.

---

## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
# Ran a one-way ANOVA with temperature_C and lakename, with aov()
temp.lake.AOV <- aov(data = FilteredLakeChemistry, temperature_C ~ lakename)
summary(temp.lake.AOV)

# Ran a one-way ANOVA with temperature_C and lakename, with lm()
temp.lake.LM <- lm(data = FilteredLakeChemistry, temperature_C ~ lakename)
summary(temp.lake.LM)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: There is a significant difference in mean temperature among the lakes (F value=50, p value=<2e-16). The same can be said for the lakes individually, with all  thusly, the null hypothesis ca



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
# Based on suggested parameters,
# created a scatter plot of temp vs depth with multiple functions.
tempVSdepth4LAKES <- FilteredLakeChemistry %>%
  ggplot(aes(x=depth,y=temperature_C,color=lakename)) +
  geom_point(alpha=0.5) + # alpha=0.5 makes the points 50% transparent
  geom_smooth(method="lm",se=FALSE) + # se=FALSE removes confidence intervals
  ylim(0,35) +
  labs(x="Depth (m)", y="Temperature (째C)", color="Lake Name") +
  custom_grid
tempVSdepth4LAKES

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
# Used TukeyHSD() to run a Tukey's HSD on the previous one-way ANOVA
TukeyHSD(temp.lake.AOV)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Lakes with the same mean temperature, statistically speaking, as Peter Lake include Paul Lake and Ward Lake. Notably, none of the lakes have a mean temperature that is statistically distinct from all the other lakes. 

DELETE BEFORE TURNING IN 1) Crampton Lake-Central Long Lake, 2) Ward Lake-Crampton Lake, 3) Hummingbird Lake-East Long Lake, 4) Tuesday Lake-East Long Lake, 5) Tuesday Lake-Hummingbird Lake, 6) West Long Lake-Hummingbird Lake, 7) Peter Lake-Paul Lake, 8) Ward Lake-Paul Lake, 9) Ward Lake-Peter Lake, and 10) West Long Lake-Tuesday Lake.

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If we were just looking at Peter Lake and Paul Lake, another test that could explore whether the two groups have distinct mean temperatures is a two-sample T-test, which tests the hypothesis that the means of two samples are equivalent.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
# Filtered the dataframe to only include records for Crampton and Ward Lakes
CramptonWardLakeChemistry <- FilteredLakeChemistry %>%
  filter(lakename=="Crampton Lake" | lakename=="Ward Lake") #instead of using & within the filter, used | to act as or

# Ran a two-sample T-test, using t.test()...
# with temperature_C as dependent variable and lakename as categorical variable
temp.lake.TTEST <- t.test(CramptonWardLakeChemistry$temperature_C ~ CramptonWardLakeChemistry$lakename)
temp.lake.TTEST

# The same test can also be formated with lm()...
# temp.lake.TTEST <- lm(CramptonWardLakeChemistry$temperature_C ~ CramptonWardLakeChemistry$lakename)
# summary(temp.lake.TTEST)

```

>Answer: Based on the results of the two-sample T-test, the Crampton and Ward Lakes statistically have the same means (t=1.1181,df=200.37,p-value=0.2649). The sample estimate for Crampton Lake is about 15.35 while the sample estimate for Ward Lake is about 14.46.
