---
title: "Assignment 2: Coding Basics"
author: "Iman Byndloss"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons/labs in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Canvas.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 55, increasing by fives. Assign this sequence a name.
```{r basics 1.1}
one_fiftyfive <- seq(1, 55, 5)
```

2.  Compute the mean and median of this sequence.
```{r basics 1.2}
mean(one_fiftyfive)
median(one_fiftyfive)
```

3.  Ask R to determine whether the mean is greater than the median.
```{r basics 1.3}
mean(one_fiftyfive) > median(one_fiftyfive)
```

4.  Insert comments in your code to describe what you are doing.

```{r basics 1.4}
#1. Creating and naming a sequence which is structured (first #, last #, by #)

#2. Using mean() to find the mean, i.e., sum divided by amount of numbers in the sequence, and median() to find the mean, i.e., the number in the middle of the sequence

#3. Using conditional statement >, which indicates greater than, to determine whether the mean is greater than the median for the named sequence
```

## Basics, Part 2

5.  Create three vectors, each with four components, consisting of (a) student names, (b) test scores, and (c) whether they are on scholarship or not (TRUE or FALSE).
```{r basics 2.1}
student <- c("Jimmy","Deshawn","Clementine","Jean") # Named and created a vector with student names, which require ""

score <- c(70, 95, 82, 99) # Named and created a vector with test scores

scholarship <- c(FALSE,TRUE,FALSE,TRUE) # Named and created a vector with scholarship information, stating either TRUE or FALSE
```

6.  Label each vector with a comment on what type of vector it is.
```{r basics 2.2}
# class() tells you what type of vector it is
class(student)

class(score)

class(scholarship)
```

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.
```{r basics 2.3}
dfSchool <- data.frame(student, score, scholarship) # Combined the vectors into a named data frame, using data.frame() and calling the vectors by name
```

8.  Label the columns of your data frame with informative titles.

```{r basics 2.4}
names(dfSchool) <- c("Student Name", "Test Score", "Scholarship") # Labeled the columns of dfSchool, using names() and listing the new labels as a vector
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: A matrix lists columns of the same length and mode. A data frame is more general than a matrix, listing vectors of equal length but different modes. 

10. Create a function with one input. In this function, use `if`...`else` to evaluate the value of the input: if it is greater than 50, print the word "Pass"; otherwise print the word "Fail". 

11. Create a second function that does the exact same thing as the previous one but uses `ifelse()` instead if `if`...`else `. 

12. Run both functions using the value 52.5 as the input

13. Run both functions using the **vector** of student test scores you created as the input. (Only one will work properly...)

```{r basics 3}
#10. Create a function using if...else
f1 <- function(x){
  if(x>50) {
    print("Pass")
  }
  else {
    print("Fail")
  }
}
# Used if() to explain what to do if expression is TRUE
# Used else() to explain what to do if expression is FALSE

#11. Create a function using ifelse()
f2 <- function(x){
  result <- ifelse(x>50, "Pass", "Fail")
  print(result)
}
# Used ifelse() which is structured (expression, if TRUE, if FALSE)
# An error occurs when print is included in the if TRUE and if FALSE statement, so it had to be removed and placed on a separate line in the function

#12a. Run the first function with the value 52.5
f1(52.5)

#12b. Run the second function with the value 52.5
f2(52.5)

#13a. Run the first function with the vector of test scores
# f1(score)

#13b. Run the second function with the vector of test scores
f2(score)
```

14. QUESTION: Which option of `if`...`else` vs. `ifelse` worked? Why? (Hint: search the web for "R vectorization")

> Answer: if...else() did not work for the vector of test scores. While ifelse(), being capable of running multiple times for different values, can handle vectorized operations, if...else() cannot handle vectorized operations due to the fact that it can only run once for a singular value.


**NOTE** Before knitting, you'll need to comment out the call to the function in Q13 that does not work. (A document can't knit if the code it contains causes an error!)